Estimating changes in temperature distributions in a large ensemble of climate simulations using quantile regression

### 2018-06-04 paper_revision.tex
my own changes before michael

With a smooth estimate of quantiles of average temperature, we show that the onset of spring, as measured by the first day of the year where the $.5$ quantile estimate reaches $-2.2^{\circ}$C \citep{pearse2017statistical}, occurs earlier in the year as the climate warms in the Detroit area (see Figure \ref{fig:FirstDayAboveFreezing} where location \textbf{b} is analyzed). The lower quantiles progress faster than the $.5$ quantile, with the $.25$ quantile hitting the $-2.2^{\circ}$C mark at a rate of approximately 15 days earlier per decade at present times. Note also that the $.5$ quantile never goes below the threshold after year 2080. It is unclear how to produce the equivalent results using existing methods of segmenting average temperature into seasons. For instance, if we were to look at quantiles of average temperature during winter the edges of the season would pull the overall quantile estimates up and prematurely estimate the onset of spring. Moreover getting information regarding the average temperature of a specific day would be impossible with seasonal averaging.

\begin{figure}[ht]
\centerline{\includegraphics[width=\smallfigwidth]{FirstDayAboveFreezing.pdf}}
\caption{\small{First day above freezing (solid lines) and $-2.2^\circ$C (dashed lines) for each year from 1850-2100 as measured by fitting quantiles to average daily temperature of the CESM ensemble data set. Three quantiles are shown to capture the spread of the distribution, $.5$ (green), $.25$ (red) and $.05$ (black).}}
\label{fig:FirstDayAboveFreezing}
\end{figure}

%To what extent the distributional characteristics are changing, and whether these changes are significant. 
Our methodology for quantile estimation provides additional information that helps to
 quantify how temperature distributions are changing and to estimate the uncertainty associated with each change. 
%The goal is to understand how the temperature distribution is changing, both in the middle and in the extremes (or tails) of the distribution. 
We can evaluate not only bulk variability  -- the interquartile range (IQR), the difference between the $0.25$ and $0.75$ quantiles -- but differences between any two quantiles. 
We therefore evaluate the difference between two low or high quantiles, denoted $\Delta q_{low}$ and $\Delta q_{high}$, which measure tail variability in the same way that interquartile range measures the variability of the bulk distribution.
If the skewness of a distribution changes over time, then future distributions are not simply scaled versions of present distributions. That is, their tail variabilities must change differently than does the IQR.
  In the case of the northern mid-latitudes winter temperatures shown in Figure \ref{meanStdSkew_Winter}, where distributions become more negatively skewed as bulk variability decreases in the future, the effect could result from either/both a low tail contracting less than the bulk (or actually increasing), or a high tail contracting more than the bulk. Our methodology allows readily differentiating these cases.

To assess whether the high tail and/or the low tail is driving changes in skewness, we consider the fractional changes in low, high, and bulk variability.
%Bulk variability, e.g. the interquartile range (IQR), is commonly linked to the tail variability. For example, a decrease in the bulk variability as a result of an overall scale change will produce a corresponding decrease in the tail variability. However, if for example the tail contracts less the bulk, there will be a skewness change. To show how variability at the different quantiles change with respect to the initial value, 
If we denote the initial and final quantile difference as $\Delta q_{q, i}$ and $\Delta q_{q, f}$ at the $q$ quantile, the temporal change in quantile differences relative to the initial year is then
\begin{equation}
	\rho = \frac{\Delta q_{q, f} - \Delta q_{q, i}}{\Delta q_{q, i}}.
\label{eq:relnorm}
\end{equation}
 Because we model the complete temperature distribution for each day of the year for all years, we choose a representative day to understand winter and summer changes (Jan 1 and July 5, respectively), and consider the difference between the beginning and end of the scenarios, the years 1850 and 2000. For these representative days, we show in Figure \ref{winter_summer_var_avg} the fractional variability changes of $\rho$ for low and high tails as well as the IQR.

Results show that tail changes can indeed differ strongly from changes in the bulk of the distribution.  In wintertime (Figure \ref{winter_summer_var_avg}, top row), in much of the northern mid-latitudes (including locations \textbf{a} and \textbf{b}), low tails change in a way that contributes to a more negative skew. Low tail variability contracts less than does the IQR, while high tail variability contracts more strongly. (High tails would contribute to more negative winter skew predomimantly in the Hudson Bay region, where the model shows distinct bias.) In summertime (Figure \ref{winter_summer_var_avg}, bottom row), the high tail dominates the transition to positive skew in the Southern Great Plains region (including location \textbf{c}).
%to the production of negative skew in southern Mexico/Central America. Bulk variability increases, but variability increases more in the low tail and less in the high tail than in the IQR.   
% See Supplementary Online Material for additional related figures that provide context.  Figure \ref{winter_summer_var_avg_big} shows the same analysis over a larger region, and Figures \ref{winter_summer_var_mn} and \ref{winter_summer_var_mx} repeat Figure \ref{winter_summer_var_avg} but for daily maximum and minimum temperatures, respectively, rather than for daily means. 
%Summer variability changes more in the low tail and the IQR than in the high tail. Substantial widening of the temperature distribution's IQR and low tail is seen in Central America, in turn causing a more negative skew as seen in Figure \ref{meanStdSkew_Summer}. Information about which tail change causes change in skewness is a detail made possible through our proposed method.

@@

While we show only three locations in the text here, an online interactive application allows similar in-depth examination of changes in model temperature distributions at all locations within North America, available at {\small\texttt{https://matzhaugen.com/links.html}}. The application allows the user to browse through any desired location to see how the variability changes as a function of season, year and quantile difference. We include temperature histograms of the first and last simulation year for the designated location, as well as maps that show the variability change spatially.

>>>>

With a smooth estimate of quantiles of average temperature, we show that the onset of spring, as measured by the first day of the year where the $.5$ quantile estimate reaches $-2.2^{\circ}$C \citep{pearse2017statistical}, occurs earlier in the year as the climate warms in the Detroit area (see Figure \ref{fig:FirstDayAboveFreezing} where location \textbf{b} is analyzed). The lower quantiles progress faster than the $.5$ quantile, with the $.25$ quantile hitting the $-2.2^{\circ}$C mark at a rate of approximately 15 days earlier per decade at present times. Note also that the $.5$ quantile never goes below the threshold after year 2080. It is unclear how to produce the equivalent results using existing methods of segmenting average temperature into seasons. For instance, if we were to look at quantiles of average temperature during winter the edges of the season would pull the overall quantile estimates up and prematurely estimate the onset of spring. Moreover, obtaining information regarding the average temperature of a specific day would be impossible with seasonal averaging.

\begin{figure}[ht]
\centerline{\includegraphics[width=\smallfigwidth]{FirstDayAboveFreezing.pdf}}
\caption{\small{First day above freezing (solid lines) and $-2.2^\circ$C (dashed lines) for each year from 1850-2100 as measured by fitting quantiles to average daily temperature of the CESM ensemble data set. Three quantiles are shown to capture the spread of the distribution, $.5$ (green), $.25$ (red) and $.05$ (black). Uncertainty, calculated from the bootstrap, was within one day for each line and thus omitted from this plot.}}
\label{fig:FirstDayAboveFreezing}
\end{figure}

%To what extent the distributional characteristics are changing, and whether these changes are significant. 
Our methodology for quantile estimation provides additional information that helps to
 quantify how temperature distributions are changing and to estimate the uncertainty associated with each change. 
%The goal is to understand how the temperature distribution is changing, both in the middle and in the extremes (or tails) of the distribution. 
We can evaluate not only bulk variability  -- the interquartile range (IQR), the difference between the $0.25$ and $0.75$ quantiles -- but differences between any two quantiles. 
We therefore evaluate the difference between two low or high quantiles, denoted $\Delta q_{low}$ and $\Delta q_{high}$, which measure tail variability in the same way that interquartile range measures the variability of the bulk distribution.
If the skewness of a distribution changes over time, then future distributions are not simply scaled versions of present distributions. That is, their tail variabilities must change differently compared with the IQR. In the case of the northern mid-latitudes winter temperatures shown in Figure \ref{meanStdSkew_Winter}, where distributions become more negatively skewed as bulk variability decreases in the future, the effect could result from either/both a low tail contracting less than the bulk (or actually increasing), or a high tail contracting more than the bulk. Our methodology allows readily differentiating these cases.

To assess whether the high tail and/or the low tail is driving changes in skewness, we consider the fractional changes in low, high, and bulk variability.
%Bulk variability, e.g. the interquartile range (IQR), is commonly linked to the tail variability. For example, a decrease in the bulk variability as a result of an overall scale change will produce a corresponding decrease in the tail variability. However, if for example the tail contracts less the bulk, there will be a skewness change. To show how variability at the different quantiles change with respect to the initial value, 
If we denote the initial and final quantile difference as $\Delta q_{q, i}$ and $\Delta q_{q, f}$ at the $q$ quantile, the temporal change in quantile differences relative to the initial year is then
\begin{equation}
	\rho = \frac{\Delta q_{q, f} - \Delta q_{q, i}}{\Delta q_{q, i}}.
\label{eq:relnorm}
\end{equation}
 Because we model the complete temperature distribution for each day of the year for all years, we choose a representative day to understand winter and summer changes (Jan 1 and July 5, respectively), and consider the difference between the beginning and end of the scenarios, the years 1850 and 2100. For these representative days, we show in Figure \ref{winter_summer_var_avg} the fractional variability changes of $\rho$ for low and high tails as well as the IQR.

Results show that tail changes can indeed differ strongly from changes in the bulk of the distribution.  In wintertime (Figure \ref{winter_summer_var_avg}, top row), in much of the northern mid-latitudes (including locations \textbf{a} and \textbf{b}), low tails change in a way that contributes to a more negative skew. Low tail variability contracts less than the IQR, while high tail variability contracts more strongly. High tails would contribute to more negative winter skew predomimantly in the Hudson Bay region, where the model shows distinct bias. In summertime (Figure \ref{winter_summer_var_avg}, bottom row), the high tail dominates the transition to positive skew in the Southern Great Plains region (including location \textbf{c}).
%to the production of negative skew in southern Mexico/Central America. Bulk variability increases, but variability increases more in the low tail and less in the high tail than in the IQR.   
% See Supplementary Online Material for additional related figures that provide context.  Figure \ref{winter_summer_var_avg_big} shows the same analysis over a larger region, and Figures \ref{winter_summer_var_mn} and \ref{winter_summer_var_mx} repeat Figure \ref{winter_summer_var_avg} but for daily maximum and minimum temperatures, respectively, rather than for daily means. 
%Summer variability changes more in the low tail and the IQR than in the high tail. Substantial widening of the temperature distribution's IQR and low tail is seen in Central America, in turn causing a more negative skew as seen in Figure \ref{meanStdSkew_Summer}. Information about which tail change causes change in skewness is a detail made possible through our proposed method.

@@

% While we show only three locations in the text here, an online interactive application allows similar in-depth examination of changes in model temperature distributions at all locations within North America, available at {\small\texttt{https://matzhaugen.com/links.html}}. The application allows the user to browse through any desired location to see how the variability changes as a function of season, year and quantile difference. We include temperature histograms of the first and last simulation year for the designated location, as well as maps that show the variability change spatially.
